\documentclass[USenglish]{uit-thesis}

\usepackage[backend=bibtex,style=numeric]{biblatex}
\usepackage{gensymb}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usepackage[toc,page]{appendix}
\usepackage{adjustbox}
\usepackage{tabularx}
\bibliography{main}
\usepackage[utf8]{inputenc}

\graphicspath{{images/}}

% Define colours
\definecolor{bblue}{HTML}{72A9DA}
\definecolor{rred}{HTML}{CA422D}
\definecolor{ggreen}{HTML}{95C380}

\pgfplotmarksize=3pt

\makeatletter
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}

\pgfplotsset{
    discontinuous/.style={
    scatter,
    scatter/@pre marker code/.code={
        \ifnodedefined{marker}{
            \pgfpointdiff{\pgfpointanchor{marker}{center}}%
             {\pgfpoint{0}{0}}%
             \ifdim\pgf@y>0pt
                \tikzset{options/.style={mark=*, fill=white}}
                \draw [densely dashed] (marker-|0,0) -- (0,0);
                \draw plot [mark=*] coordinates {(marker-|0,0)};
             \else
                \tikzset{options/.style={mark=none}}
             \fi
        }{
            \tikzset{options/.style={mark=none}}
        }
        \coordinate (marker) at (0,0);
        \begin{scope}[options]
    },
    scatter/@post marker code/.code={\end{scope}}
    }
}

\makeatother

\begin{document}

%:-------------------------- Frontpage ------------------------

\title{GeneNet VR: Large Biological Networks in Virtual Reality Using Inexpensive Hardware}
% \subtitle{Subtitle}% Note: this is optional, and may be commented out
\author{\'Alvaro Mart\'inez Fern\'andez}
\thesisfaculty{Faculty of Science and Technology \\ Department of Computer Science}
\thesisprogramme{INF-3990 Master's thesis in Computer Science November 2020}

\pgfplotsset{width=\textwidth}

\maketitle

%:-------------------------- Frontmatter -----------------------
\frontmatter

%: Commented out
% \iffalse
% \begin{dedication}
% To...
%
% Thanks for...
% \end{dedication}

\begin{epigraph}
\epigraphitem{Anything created by human beings is already in the great book of nature.}{Antoni Gaudi}
\epigraphitem{There are decades of innovations ahead. We’re at the very beginning, where it’s just at the stage where we can bring in consumers [but] there’s so much further to go from there.}{Brendan Iribe}
\end{epigraph}

\begin{abstract}
The visualization of large amounts of biological data can be challenging. Network representations are usually preferable for certain kinds of biological data, but they face problems such as information overload, high interconnectivity and high dimensionality. Some visualization tools try to solve these problems by reducing the interactivity in favour of more information or by using specific hardware. This thesis aims to improve these problems using Virtual Reality (VR) and the Oculus Quest, an affordable VR headset.

In order to test our hypothesis that Virtual Reality can be advantageous in the visualization of large biological networks, we built GeneNet VR, a prototype for the Oculus Quest. We created a case study where we used two gene networks from MIxT, a real application with visualization issues. A performance and scalability evaluation was carried out in addition to a qualitative research approach where we conducted purposive sampling through in-depth semi-structured interviews with several research scientists from UiT. The results showed that the performance for the datasets from our use case reaches the FPS required by Oculus and that GeneNet VR scales well for those network sizes. Oculus Quest is therefore an affordable solution for the visualization of large datasets. From the interviews we also learned that GeneNet VR has potential as a visualization tool and that VR demonstrates benefits for pattern finding in complex networks. GeneNet VR has also shown to be easy to use even for novice VR users. Additionally, we gathered interesting feedback and knowledge for further improvements applicable not only for biological networks, but also in other fields like drug and social networks.

From the results we can say that Virtual Reality has the capability to solve common problems in the visualization of large biological networks. The rich interactivity that VR offers can compensate for the large amounts of data. With the advancement in hardware, VR headsets have also become more resourceful and economical. Thus, VR hardware like the Oculus Quest, should be considered a competitive solution for visualization tools, as described in this thesis. 

\end{abstract}

\begin{acknowledgement}

I would first like to thank my advisor, Associate Professor Edvard Pedersen, for his guidance through this thesis, his constant encouragement and for the interesting discussions about Virtual Reality and Computer Science. I would also like to thank my co-advisor, Lars Ailo Bongo, for his great knowledge in Bioinformatics, his good criticism and for all the great insights about scientific writing. Thank you also Vanessa Dumeaux, for introducing me to MIxT and for your great contributions in Biolology.

I would also like to thank the collaboration of the several research scientists that participated in the evaluation of my project and that contributed with their knowledge in Biology and Computer Science.

To my close friends Juncal García García, Mireia Nager and Reidar Staupe-Delgado for all the great experiences in Tromsø and for teaching me the value of science.

To Ramsalt Lab, for making it possible for me to study and expand my knowledge in Computer Science.

I would like to thank also my family, for their support and encouragement even from the distance.

Finally, thank you Dominic Ochotorena for your love, for always supporting me and for being there.

\end{acknowledgement}
% \fi
%: End comments

\tableofcontents

%:-------------------------- Mainmatter -----------------------
\mainmatter

\chapter{Introduction}
\input{chapters/introduction}

\chapter{GeneNet VR}
\input{chapters/genenetvr}

\chapter{MIxT}
\input{chapters/mixt}

\chapter{Related work}
\input{chapters/related-work}

\chapter{Evaluation and discussion}
\input{chapters/evaluation}

\chapter{Conclusion}
\input{chapters/conclusion}

\chapter{Future work}
\input{chapters/futureWork}

\printbibliography

% \begin{appendices}
% \chapter{Appendix A}
% \input{chapters/appendixBiopsy}
% \end{appendices}

\backmatter

\end{document}
